---
title: "Accessory_genome_analysis"
format: html
editor: visual
---

# Accessory_genome_analysis

## All 139 isolates

Whole genome sequence data from 139 STEC O157 were included in the below analysis. This is based off Dr Sara Burgess's "AMR dairy calf project - modelling analysis". A gene presence absence matrix was generated using Panaroo, and the core and accessory genomes were extracted from the gene_presence_absence.csv, with data tied along the way based on what was needed for the functions.

https://link.springer.com/article/10.1186/s12915-023-01562-w#Sec11 https://journals.asm.org/doi/10.1128/spectrum.00956-25 https://link.springer.com/article/10.1007/s10096-023-04600-1#Sec2 https://pmc.ncbi.nlm.nih.gov/articles/PMC12772303/#s2

```{r libraries, message = FALSE, warning = FALSE}
# Load libraries
library(vegan)
library(BiocManager)
library(dplyr)
library(tidyr)
library(gtools)
library(effects)
library(sjstats)
library(lme4)
library(dbstats)
library(ggplot2)
library(glue)
library(randomForest)
library(car)
library(performance)
```

### Seperating core and accessory genomes

The panaroo default for the core genome is that 95% of the isolates must contain the gene therefore to be included in the core genome 132 of the genomes must have a gene

```{r panaroo_core_accessory, results = 'hide'}
setwd("~/STEC_139/Objective_2_pangenome_analysis")

# lets try combining column 1 and 3 together as column 1 is all unique, but column 3 gives the actual annotation. 

# Load Panaroo matrix
gene_matrix <- read.csv("~/STEC_139/Objective_2_pangenome_analysis/Data/gene_presence_absence_panaroo.csv", stringsAsFactors = FALSE,  # prevent factor conversion
  check.names = FALSE       # prevents R from changing isolate names
)

# combining columns 1 and 3 making new variable called full_name
gene_matrix$full_name<-paste(gene_matrix$Gene, gene_matrix$Annotation)

gene_reordered<-gene_matrix[c("full_name", setdiff(names(gene_matrix), "full_name"))]
print(gene_reordered)

# Save it to make things easier
write.csv(gene_reordered, "~/STEC_139/Objective_2_pangenome_analysis/Data/gene_reordered.csv", row.names = FALSE)

# ok, lets load in the new matrix
gene_reordered_matrix <- read.csv(
  "~/STEC_139/Objective_2_pangenome_analysis/Data/gene_reordered.csv",
  row.names = 1,       # first column is gene names
  stringsAsFactors = FALSE,  # prevent factor conversion
  check.names = FALSE       # prevents R from changing isolate names
)

# the isolate names are werid, so lets sort those out
colnames(gene_reordered_matrix) <- sub("^fix_", "", colnames(gene_reordered_matrix))

# Id gene_matrix identifing isolate columns
isolate_cols <- grep("^EcCa", colnames(gene_reordered_matrix))

# Convert isolate columns to binary 0/1
gene_reordered_matrix_binary <- gene_reordered_matrix %>%
  mutate(across(all_of(isolate_cols),
                ~ ifelse(.x == "" | is.na(.x), 0, 1)))

# Add TotalCount = number of isolates where gene is present
gene_reordered_matrix_binary <- gene_reordered_matrix_binary %>%
  mutate(TotalCount = rowSums(across(all_of(isolate_cols))))

#The panaroo default for the core genome is that 95% of the isolates must contain the gene therefore to be included in the core genome 132 of the genomes must have a gene
139*0.95

########
##Core##
########

#Selecting those rows that are part of the core genome 
gene_reordered_matrix_core<- gene_reordered_matrix_binary %>%
  filter(TotalCount >= 132)

summary(gene_reordered_matrix_binary$TotalCount)

#Removing the the TotalCount column
gene_reordered_matrix_core <- gene_reordered_matrix_core %>%
  select(-TotalCount)

#Remove "_medaka" and "fix" from header names - this was done above
#gene_reordered_matrix_core <- gene_reordered_matrix_core%>%
 # rename_with(~gsub("_medaka$", "", .), ends_with("_medaka"))

colnames(gene_reordered_matrix_core) <- sub("^fix_", "", colnames(gene_reordered_matrix_core))

# Re-identify isolate columns (names have changed)
isolate_cols_core <- grep("^EcCa", colnames(gene_reordered_matrix_core))

# Subset to isolate-only matrix
core_binary_only <- gene_reordered_matrix_core[, isolate_cols_core]

# Transpose
transposed_core <- t(core_binary_only)

# Reorder isolates to match metadata
transposed_core <- transposed_core[order(rownames(transposed_core)), ]

# Compute the distance matrix
dist_matrix_core <- vegdist(transposed_core, method = "jaccard", binary = TRUE)


#############
##Accessory##
#############

#Selecting those rows that are part of the accessory genome
gene_reordered_matrix_accessory <- gene_reordered_matrix_binary %>%
  filter(TotalCount < 138)

#Remove the TotalCount column
gene_reordered_matrix_accessory <- gene_reordered_matrix_accessory %>%
  select(-TotalCount)

#Remove "_medaka" from header names
gene_reordered_matrix_accessory <- gene_reordered_matrix_accessory%>%
  rename_with(~gsub("_medaka$", "", .), ends_with("_medaka"))

# Re-identify isolate columns (names have changed)
isolate_cols_accessory <- grep("^EcCa", colnames(gene_reordered_matrix_accessory))

# Subset to isolate-only matrix
accessory_binary_only <- gene_reordered_matrix_accessory[, isolate_cols_accessory]

#Transpose table so that columns become rows
transposed_accessory <- t(accessory_binary_only)

# Reorder isolates to match metadata
transposed_accessory  <- transposed_accessory [order(rownames(transposed_accessory )), ]


################
# Convert to dataframe if not already
transposed_accessory_df <- as.data.frame(transposed_accessory)

# Optional: add rownames as a column so editable
transposed_accessory_df$Isolate_ID <- rownames(transposed_accessory_df)

# Reorder so Isolate_ID is first column
transposed_accessory_df <- transposed_accessory_df %>%
  select(Isolate_ID, everything())

# Export to CSV
write.csv(transposed_accessory_df, "~/STEC_139/Objective_2_pangenome_analysis/Data/transposed_accessory_for_excel.csv", row.names = FALSE)

############################
## Double checking things ##
############################


# Number of isolates (columns before transpose)
ncol(accessory_binary_only)
# Or after transpose (rows) - lets just double check please
nrow(transposed_accessory)


# Original isolates
all_isolates <- colnames(core_binary_only)  # original gene_reordered_matrix_binary columns
# Isolates in accessory
accessory_isolates <- colnames(accessory_binary_only)

# Check which isolates are missing
missing_in_accessory <- setdiff(all_isolates, accessory_isolates)
missing_in_accessory
```

```{r accessory_data_setup}
# Load in metadata, isolate IDs should be matching now
metadata <- read.csv("~/STEC_139/Objective_2_pangenome_analysis/Data/metadata_Apr26.csv", header = TRUE, stringsAsFactors = FALSE)
# The transposed 
transposed_accessory_for_excel <- read.csv("~/STEC_139/Objective_2_pangenome_analysis/Data/transposed_accessory_for_excel.csv", stringsAsFactors=TRUE)

# compute distance matrix
dist_matrix_accessory <- vegdist(transposed_accessory_for_excel %>% select(-Isolate_ID),
                                 method = "jaccard", binary = TRUE)
```

### PCoA

```{r PCoA_accessory_analysis}
# Number of isolates in distance matrix
length(attr(dist_matrix_accessory, "Labels"))

# Classical multidimensional scaling (PCoA)
accessory_pcoa <- cmdscale(dist_matrix_accessory, eig = TRUE, k = 2)  # k = 2 for 2D

# Make dataframe of PCoA coordinates
pcoa_coords <- as.data.frame(accessory_pcoa$points)
colnames(pcoa_coords) <- c("PCoA1", "PCoA2")

# Add isolate IDs (Excel-corrected dataframe)
pcoa_coords$Isolate_ID <- transposed_accessory_for_excel$Isolate_ID

# Quick check
head(pcoa_coords)

# Ensure the metadata has the same Isolate_ID column
pcoa_plot_data <- pcoa_coords %>%
  left_join(metadata, by = "Isolate_ID")

head(pcoa_plot_data)


# Eigenvalues from PCoA
eig_values <- accessory_pcoa$eig

# Calculate proportion of variance for each axis
variance_explained <- round(100 * eig_values / sum(eig_values), 1)

# Variance explained by PCoA1 and PCoA2
variance_explained[1:2]

##############
## Plotting ##
##############

# SBI
ggplot(pcoa_plot_data, aes(x = PCoA1, y = PCoA2, color = SBI)) +
  geom_jitter(width = 0.03, height = 0.03, size = 3, alpha = 0.8)  +
  stat_ellipse(aes(fill = SBI), geom = "polygon", alpha = 0.2) +
  theme_minimal() +
  labs(
    title = "Accessory Genome PCoA",
    x = paste0("PCoA1 (", variance_explained[1], "%)"),
    y = paste0("PCoA2 (", variance_explained[2], "%)")
  )

# sample site
ggplot(pcoa_plot_data, aes(x = PCoA1, y = PCoA2, color = Sample_site)) +
  geom_jitter(width = 0.05, height = 0.05, size = 3, alpha = 0.8) +
  stat_ellipse(aes(fill = Sample_site), geom = "polygon", alpha = 0.2) +
  theme_minimal() +
  labs(
    title = "Accessory Genome PCoA",
    x = paste0("PCoA1 (", variance_explained[1], "%)"),
    y = paste0("PCoA2 (", variance_explained[2], "%)")
  )
```

### Significant genes

```{r signif_genes}

##################
## Signif genes ##
##################

accessory_binary <- transposed_accessory_for_excel %>%
  select(-Isolate_ID)

# Ensure numeric 0/1
accessory_binary[] <- lapply(accessory_binary, function(x) as.numeric(x > 0))

# Remove zero-variance genes
accessory_binary_var <- accessory_binary[, apply(accessory_binary, 2, sd) > 0]

# PC1 signif genes

pc1_gene_cor <- apply(
  accessory_binary_var,
  2, function(x) cor(x, accessory_pcoa$points[, 1]))

pc1_df <- data.frame(
  Gene = names(pc1_gene_cor),
  Correlation_with_PCoA1 = pc1_gene_cor) %>% arrange(desc(abs(Correlation_with_PCoA1)))

head(pc1_df, 10)

# PC2 signif genes

pc2_gene_cor <- apply(
  accessory_binary_var,
  2, function(x) cor(x, accessory_pcoa$points[, 2]))

pc2_df <- data.frame(
  Gene = names(pc2_gene_cor),
  Correlation_with_PCoA2 = pc2_gene_cor) %>% arrange(desc(abs(Correlation_with_PCoA2)))

head(pc2_df, 10)

# Lets save the files for easy access

write.csv(pc1_df, "~/STEC_139/Objective_2_pangenome_analysis/Data/pc1_df.csv", row.names = FALSE)
write.csv(pc2_df, "~/STEC_139/Objective_2_pangenome_analysis/Data/pc2_df.csv", row.names = FALSE)
```

### Stx genes

```{r stx_genes}
###############
## stx genes ##
###############

# Find stx in a case insensitive manner
stx_genes <- grep("stx", rownames(pc1_df), ignore.case = TRUE, value = TRUE)

selected_data_stx_genes <- pc1_df[stx_genes, ]

print(selected_data_stx_genes)
```

### BETA diversity

This is the PERMANOVA and PERMdisp tests. It measures the change or turnover of species composition across different ecosystems, habitats, or geographic regions.

```{r permanova_accessory}
####################
## beta diversity ##
####################

#Permanova of accessory genes
set.seed(312)
perm_accessory <- adonis2(dist_matrix_accessory ~ Sample_site , data = metadata, permutations = 1000, by = "margin")

perm_accessory

```

## Farm 84 isolates

```{r keeping_farm_84}
##############
## farm 84 ###
##############

str(transposed_accessory_for_excel)

# Lets list the isolates individually, just so I make sure I have them all
keep_isolates <- c(
"EcCa531a","EcCa531b","EcCa531c","EcCa531d",
"EcCa532a","EcCa533a",
"EcCa534a","EcCa534b","EcCa534c","EcCa534d",
"EcCa535a","EcCa535b","EcCa535c","EcCa535d",
"EcCa536a","EcCa536b","EcCa536c","EcCa536d","EcCa536e",
"EcCa539a",
"EcCa540a","EcCa540b",
"EcCa541a","EcCa541b","EcCa541c",
"EcCa542a","EcCa542b","EcCa542c",
"EcCa543a","EcCa543b","EcCa543c","EcCa543d",
"EcCa626a","EcCa626b","EcCa626c","EcCa626d",
"EcCa628a","EcCa629a_08",
"EcCa630a","EcCa630b","EcCa630c","EcCa630d",
"EcCa631a","EcCa631b",
"EcCa632a","EcCa632b",
"EcCa633a","EcCa633b","EcCa633c","EcCa633d",
"EcCa637a",
"EcCa645a","EcCa645b","EcCa645c",
"EcCa646a","EcCa646b","EcCa646c","EcCa646d_comb",
"EcCa647a","EcCa647b","EcCa647c","EcCa647d"
)

# check we have them all
length(keep_isolates)

# make dataframe
transposed_accessory_filtered <- transposed_accessory_for_excel[
  transposed_accessory_for_excel$Isolate_ID %in% keep_isolates, 
]

# more checking
nrow(transposed_accessory_filtered)

setdiff(keep_isolates, rownames(transposed_accessory_for_excel))

```

```{r metadata_farm84}
# load in the corresponding metadata
metadata_filtered <- read.csv("~/STEC_139/Objective_2_pangenome_analysis/Data/metadata_filtered.csv", header = TRUE, stringsAsFactors = FALSE)

# compute distance matrix
dist_matrix_accessory_filtered <- vegdist(transposed_accessory_filtered %>% select(-Isolate_ID),
                                 method = "jaccard", binary = TRUE)
```


|   Concept   	|             Alpha diversity            	|        Beta diversity       	|
|:-----------:	|:--------------------------------------:	|:---------------------------:	|
| Level       	| Within isolate                         	| Between isolates            	|
| Question    	| How many genes does each isolate have? 	| How different are isolates? 	|
| Your metric 	| Richness, Shannon, Simpson             	| Jaccard distance            	|
| Output      	| Boxplots, summaries                    	| PCoA, NMDS, clustering      	|


#### ALPHA diversity

Shannon and Simpson diversity tests.

Alpha diversity ((\alpha )-diversity) is the mean species diversity within a particular area, habitat, or ecosystem at a local scale. It measures the diversity of a single community and is typically assessed by evaluating two core components: species richness (the total number of different species) and species evenness (how evenly the individuals are distributed among those species).

```{r alphadiv_farm84}
#####################
## alpha diversity ##
#####################

library(vegan)
library(dplyr)

# Remove isolate ID column
accessory_matrix <- transposed_accessory_filtered %>%
  select(-Isolate_ID)

# Ensure all columns are numeric
accessory_matrix[] <- lapply(accessory_matrix, as.numeric)

# Check structure
str(accessory_matrix)

# Richness (the number of accessory genes present)
richness <- rowSums(accessory_matrix)
#specnumber(accessory_matrix)

#specnumber(accessory_matrix, groups = Sample_site)

# Shannon diversity
shannon_div <- diversity(accessory_matrix, index = "shannon")

# Simpson diversity
simpson_div <- diversity(accessory_matrix, index = "simpson")

# Combine into dataframe
diversity_results <- data.frame(
  Isolate_ID = transposed_accessory_filtered$Isolate_ID,
  Richness = richness,
  Shannon = shannon_div,
  Simpson = simpson_div
)

head(diversity_results)

diversity_results <- diversity_results %>%
  left_join(metadata_filtered, by = "Isolate_ID")

##############
## Plotting ##
##############

Sample_site_colours <- c(
  "Farm"     = "#88ccee",
  "Lairage" = "#44aa99",
  "Hide"  = "#DDCC77",
  "Pre_intervention" = "#cc6677",
  "Post_intervention" = "#332288"
)


P_rich_84 <- ggplot(diversity_results,
       aes(x = Sample_site,
           y = Richness,
           fill = Sample_site)) +
  geom_boxplot() +
   scale_fill_manual(values = Sample_site_colours) +
  scale_x_discrete(limits = c("Farm", "Lairage", "Hide", "Pre_intervention", "Post_intervention")) +
  labs(y = "Accessory gene richness (Counts of genes)") +
  theme_minimal()

P_rich_84

ggsave("~/STEC_139/Objective_2_pangenome_analysis/Figures/P_rich_84.svg", plot = P_rich_84, width = 7, height = 5)

ggplot(diversity_results,
       aes(x = Sample_site,
           y = Shannon,
           fill = Sample_site)) +
  geom_boxplot() +
  theme_minimal()

ggplot(diversity_results,
       aes(x = Sample_site,
           y = Simpson,
           fill = Sample_site)) +
  geom_boxplot() +
  theme_minimal()

```

| Isolate | Gene1 | Gene2 | Gene3 | Gene4 | Richness |
|:-------:|:-----:|:-----:|:-----:|:-----:|:--------:|
| EcCa001 | 1     | 1     | 0     | 1     | 3        |
| EcCa002 | 1     | 0     | 0     | 1     | 2        |

For example the richness is the total number of genes per isolate. 

EcCa001 = 3 accessory genes present
EcCa002 = 2 accessory genes present

So in the context of an accessory genome presence/absence matrix:

Richness = number of accessory genes present
Shannon diversity = incorporates both the number of genes and their relative abundances
Simpson diversity = measures dominance/evenness of genes

The boxplot then shows the richness per site in terms of the accessory genes of the isolates of that specific site, It is purely showing the difference in numbers not the variety or difference of the genes. This is looking at how many genes isolates contain. 

```{r alphadiv_samplesite_farm84}
diversity_summary <- diversity_results %>%
  group_by(Sample_site) %>%
  summarise(
    n = n(),

    Mean_Richness = round(mean(Richness), 2),
    SD_Richness = round(sd(Richness), 2),

    Mean_Shannon = round(mean(Shannon), 2),
    SD_Shannon = round(sd(Shannon), 2),

    Mean_Simpson = round(mean(Simpson), 2),
    SD_Simpson = round(sd(Simpson), 2)
  )

diversity_summary

kruskal.test(Richness ~ Sample_site,
             data = diversity_results)

kruskal.test(Shannon ~ Sample_site,
             data = diversity_results)

kruskal.test(Simpson ~ Sample_site,
             data = diversity_results)

cor(diversity_results$Richness,
    diversity_results$Shannon)

cor(diversity_results$Richness,
    diversity_results$Simpson)
```

	Kruskal-Wallis rank sum test

data:  Richness by Sample_site
Kruskal-Wallis chi-squared = 8.0652, df = 4, p-value = 0.08922

H0: The richness distributions are the same across all sample sites.
H1: The richness distributions are different across all sample sites.
One overall test - omnibus test - so not multiple testing, but that means no pairwise comparisons. 

P-value low the null must go (< 0.05). The p-value is not below the significance level of less than 0.05, so cannot reject the null hypothesis. Therefore the richness/variation of accessory genes does not differ significantly between sample sites. Accessory gene richness does not differ significantly among sample sites.

chi-squared
It measures how different the ranks are among groups.
Small value - groups have similar ranks.
Large value - groups have different ranks.

Degrees of freedom 
number of groups - 1

This test does not account for multiple testing like the dunn test does. This is because only one hypothesis is preformed is there any difference amoung the groups. It does not control for the false discovery rate. 


``` {r do_they_really_not_differ}
library(FSA)

dunnTest(Richness ~ Sample_site,
         data = diversity_results,
         method = "bh")
```
H0: The distributions of the two groups are the same.

Which specific groups differ? 

Unadj p-values are not accounting for the presence of the other sample sites. This can be type 1 error. 
The adjusted p-value accounts for the multiple testing. The fact that many comparisons are happening at once. The method = controls the false discovery rate. 

Kruskal–Wallis:
“Do sample sites differ in richness overall?”
Dunn’s test:
“Which sample sites differ in richness?”
Only Dunn’s test needs adjusted p-values (BH, etc.).

Warning :Sample_site was coerced to a factor.
  Kruskal-Wallis rank sum test

data: x and g
Kruskal-Wallis chi-squared = 8.0652, df = 4, p-value = 0.09


                      Dunn's Pairwise Comparison of x by g                      
                              (Benjamini-Hochberg)                              

Col Mean-│
Row Mean │       Farm       Hide    Lairage   Post_int
─────────┼────────────────────────────────────────────
    Hide │   0.081099
         │     1.0000 
         │
 Lairage │   2.453916   2.380814
         │     0.1413     0.0864 
         │
Post_int │   0.985162   0.937332  -0.596914
         │     0.6491     0.5810     0.6882 
         │
Pre_inte │   1.138260   1.078749  -0.810533  -0.061720
         │     0.8500     0.7017     0.5966     0.9508 

FDR = 0.05
Reject Ho if adjusted p ≤ FDR with stopping rule, where (unadjusted) p = Pr(|Z| ≥ |z|)



FDR = false discovery rate

### BETA diversity


### PCoA

```{r PCoA_farm84}
# Number of isolates in distance matrix
length(attr(dist_matrix_accessory_filtered, "Labels"))

# Classical multidimensional scaling (PCoA)
accessory_pcoa_filtered <- cmdscale(dist_matrix_accessory_filtered, eig = TRUE, k = 2)

# Make dataframe of PCoA coordinates
pcoa_coords_filtered <- as.data.frame(accessory_pcoa_filtered$points)
colnames(pcoa_coords_filtered) <- c("PCoA1", "PCoA2")

# Add isolate IDs (must match order of PCoA output)
pcoa_coords_filtered$Isolate_ID <- transposed_accessory_filtered$Isolate_ID


# Quick check
head(pcoa_coords_filtered)


# Join metadata
pcoa_plot_data_filtered <- pcoa_coords_filtered %>%
  left_join(metadata_filtered, by = "Isolate_ID")


head(pcoa_plot_data_filtered)


# Eigenvalues from PCoA
eig_values_filtered <- accessory_pcoa_filtered$eig

# Calculate proportion of variance
variance_explained_filtered <- round(100 * eig_values_filtered / sum(eig_values_filtered), 1)

# Variance explained by first two axes
variance_explained_filtered[1:2]

##############
## Plotting ##
##############

# setting SBI colours
SBI_colours <- c(
  "AWY2a"     = "#2CA02C",
  "AY2a" = "#1F77B4",
  "SY2c"  = "#FFBB78"
)


# Plot by SBI
P_1_84<-ggplot(pcoa_plot_data_filtered, aes(x = PCoA1, y = PCoA2, color = SBI)) +
  geom_jitter(width = 0.05, height = 0.05, size = 3, alpha = 0.8) +
  stat_ellipse(aes(fill = SBI), geom = "polygon", alpha = 0.2, type = "norm") +
  scale_color_manual(values = SBI_colours) +
  scale_fill_manual(values = SBI_colours) +
  theme_minimal() +
  labs(
    x = paste0("Axis.1 [", variance_explained_filtered[1], "%]"),
    y = paste0("Axis.2 [", variance_explained_filtered[2], "%]")
  )

P_1_84

Sample_site_colours <- c(
  "Farm"     = "#88ccee",
  "Lairage" = "#44aa99",
  "Hide"  = "#DDCC77",
  "Pre_intervention" = "#cc6677",
  "Post_intervention" = "#332288"
)

ggsave("~/STEC_139/Objective_2_pangenome_analysis/Figures/P_1_84.svg", plot = P_1_84, width = 7, height = 5)

# Plot by Sample_site
P_2_84<-ggplot(pcoa_plot_data_filtered, aes(x = PCoA1, y = PCoA2, color = Sample_site)) +
  geom_jitter(width = 0.15, height = 0.15, size = 3, alpha = 0.8) +
  stat_ellipse(aes(fill = Sample_site), geom = "polygon", alpha = 0.2, type = "norm") +
  scale_color_manual(values = Sample_site_colours) +
  scale_fill_manual(values = Sample_site_colours) +
  theme_minimal() +
  labs(
    x = paste0("Axis.1 [", variance_explained_filtered[1], "%]"),
    y = paste0("Axis.2 [", variance_explained_filtered[2], "%]")
  )

P_2_84

ggsave("~/STEC_139/Objective_2_pangenome_analysis/Figures/P_2_84.svg", plot =P_2_84, width = 7, height = 5)
```

### Significant genes

```{r signif_genes_farm84}
# Create binary accessory matrix (filtered)
accessory_binary_filtered <- transposed_accessory_filtered %>%
  select(-Isolate_ID)

# Ensure numeric 0/1
accessory_binary_filtered[] <- lapply(
  accessory_binary_filtered,
  function(x) as.numeric(x > 0)
)

# Remove zero-variance genes
accessory_binary_var_filtered <- accessory_binary_filtered[
  , apply(accessory_binary_filtered, 2, sd) > 0
]

##################
## Signif genes ##
##################

# PC1 signif genes
pc1_gene_cor_filtered <- apply(
  accessory_binary_var_filtered,
  2,
  function(x) cor(x, accessory_pcoa_filtered$points[, 1])
)

pc1_df_filtered <- data.frame(
  Gene = names(pc1_gene_cor_filtered),
  Correlation_with_PCoA1 = pc1_gene_cor_filtered
) %>%
  arrange(desc(abs(Correlation_with_PCoA1)))

head(pc1_df_filtered, 10)

# PC2 signif genes
pc2_gene_cor_filtered <- apply(
  accessory_binary_var_filtered,
  2,
  function(x) cor(x, accessory_pcoa_filtered$points[, 2])
)

pc2_df_filtered <- data.frame(
  Gene = names(pc2_gene_cor_filtered),
  Correlation_with_PCoA2 = pc2_gene_cor_filtered
) %>%
  arrange(desc(abs(Correlation_with_PCoA2)))

head(pc2_df_filtered, 10)


# Lets save the files for easy access
write.csv(pc1_df_filtered, "~/STEC_139/Objective_2_pangenome_analysis/Data/pc1_df_filtered.csv", row.names = FALSE)
write.csv(pc2_df_filtered, "~/STEC_139/Objective_2_pangenome_analysis/Data/pc2_df_filtered.csv", row.names = FALSE)

```

### Stx genes

```{r stx_genes_farm84}
###############
## stx genes ##
###############

# where are they? 
# Find stx in a case insensitive manner
stx_genes_filtered <- grep("stx", rownames(pc1_df_filtered), ignore.case = TRUE, value = TRUE)

selected_data_stx_genes_filtered <- pc1_df[stx_genes, ]

print(selected_data_stx_genes_filtered)
```



This is the PERMANOVA and PERMdisp tests. It measures the change or turnover of species composition across different ecosystems, habitats, or geographic regions.

```{r permanova_farm84}
####################
## beta diversity ##
####################

#Permanova of accessory genes
set.seed(312)
perm_accessory <- adonis2(dist_matrix_accessory_filtered  ~ Sample_site , data = metadata_filtered, permutations = 1000, by = "margin")

perm_accessory

permanova_table <- as.data.frame(perm_accessory)
write.csv(permanova_table, file = "~/STEC_139/Objective_2_pangenome_analysis/Tables/permanova_results_farm84.csv", row.names = TRUE)
```

|             | Df | SumOfSqs | R2       | F        | Pr(>F)      |
|-------------|----|----------|----------|----------|-------------|
| Sample_site | 4  | 0.909969 | 0.231993 | 4.304511 | 0.003996*** |
| Residual    | 57 | 3.012435 | 0.768007 |          |             |
| Total       | 61 | 3.922404 | 1        |          |             |

The permanova test results in a p-value of 0.004 (rounded to 3 decimal points), this shows that the sample site is linked to predicting the genes present in the accessory genome.

### PermDISP

```{r permdisp}
# already have a distance matrix "dist_matrix_accessory_filtered" and jaccard distance was used 
# reminder it is "metadata_filtered" to use

disp <- betadisper(dist_matrix_accessory_filtered,
                   group = metadata_filtered$Sample_site,
                   type = "centroid")

permutest(disp)

TukeyHSD(disp)
```
Permdisp interpretation: 
Do some sample sites contain more diverse accessory genomes than others?
H0: The multivariate dispersion (variability) is equal among groups.

Tukey's test interpretation:
Positive diff = first site is more dispersed.
Negative diff = second site is more dispersed.
Significant adjusted p-values indicate different dispersions.

PERMANOVA significant, PERMDISP not significant - evidence that group centroids differ (usually the ideal situation).
PERMANOVA significant, PERMDISP significant - some of the PERMANOVA signal could be due to differences in dispersion (spread) among sites, so interpretation requires more caution.
PERMDISP significant + TukeyHSD significant - identifies which sites differ in accessory genome variability.

